---
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
bibliography: "`r file.path(system.file(package='alevin2bioc', 'vignettes'), 'library.bib')`"
vignette: >
  %\VignetteIndexEntry{alevin2bioc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---

# Importing alevin scRNA-seq counts into R/Bioconductor

## Instructor(s) name(s) and contact information

[Michael Love](https://mikelove.github.io),
[Avi Srivastava](https://k3yavi.github.io)


```{r setup, echo=FALSE, results="hide"}
knitr::opts_chunk$set(cache=TRUE)
```

## Introduction

*alevin* is a method for ... [@alevin]. It extends the methods in the
*Salmon* software [@salmon].

The data we will use is ... (citation).

## Running alevin

In order to run *alevin*, we must first ...

Instructions on downloading *alevin* can be found here ...

Instructions on indexing a set of reference transcripts can be found
here ... For this experiment, we used the 
[GENCODE](https://www.gencodegenes.org/) 
human reference transcripts [@gencode].

## Importing alevin data with tximeta

We will use *tximeta* to import the *alevin* counts into
R/Bioconductor. The main function `tximeta` reads information from the
entire output directory of *alevin* or *Salmon* in order to
automatically detect and download metadata about the reference
sequences (the transcripts) [@tximeta]. It should work "out of the
box" for human, mouse, and fruit fly reference transcripts from
GENCODE, Ensembl, or RefSeq

First we specify the path where the quantification data is stored. In
this tutorial, the data is stored in an R package, so we need to use
the `system.file` command. 

**Note:** For typical use, you would not use `system.file`, but would
just specify the path to the directory of the output from *alevin*.

**Second note:** when sharing *alevin* or *Salmon* datasets, make sure
to share the entire output directory (you can `zip` or `tar` the
directories to make them into a single share-able file).

```{r}
# normally you just would use:
# dir <- "/path/to/alevin/output"
extdata <- system.file("extdata", package="alevin2bioc")
dir <- file.path(extdata, "pbmc_1k")
```

We make a vector called `files` (for *alevin* this will point to just
a single file, but with *Salmon* usually one would import multiple
files at a time).

```{r}
files <- file.path(dir, "alevin", "quants_mat.gz")
file.exists(files)
```

```{r echo=FALSE}
suppressPackageStartupMessages(library(GenomicFeatures))
```

We can import the *alevin* quantification using the following call to
`tximeta`. The extra argument to *alevin* will filter out cells based
on *alevin*'s post-quantification quality control methods (see paper
for details).

```{r}
library(tximeta)
se <- tximeta(files, type="alevin", alevinArgs=list(filterBarcodes=TRUE))
```

We can see as it was importing that it automatically detected we are
working with data quantified using GENCODE's *Homo sapiens* reference
transcripts, release 33. The transcript metadata was added
programmatically to the output (this object, `se`). `tximeta` is also
written in such a way that it will avoid unnecessary downloads and
parsing of files -- it will first attempt to find a cached version of
the metadata, to avoid re-downloading or re-parsing these files.

`tximeta` returns a *SummarizedExperiment* [@Lawrence2013]. We can
easily convert this object into a *SingleCellExperiment*
[@Amezquita2020] which has specific slots designed for single-cell
experiment data.

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
sce <- as(se, "SingleCellExperiment")
```

The *SingleCellExperiment* object is used widely across Bioconductor
packages (as you may already know), and so the code up to this point
can be used as an entry point into other Bioconductor single cell
workflows. For more details on working with *SingleCellExperiment*
objects, one can consult the following online book: 
[Orchestrating Single-Cell Analysis with Bioconductor](https://osca.bioconductor.org/)
[@Amezquita2020].

## Benefits of tximeta

We can automatically add IDs, because *tximeta* knows the type of
identifiers on the rows of the `sce` object:

```{r}
library(org.Hs.eg.db)
sce <- addIds(sce, "SYMBOL")
```

```{r}
mcols(sce)
```

Also, because the provenance was detected, we also have the ranges of
the genes in their proper genomic context. So it is easy to find, for
example, genes near a particular position in the genome, in this case
4 genes that overlap the range `chr1:10,000,000-10,100,000`.

```{r}
x <- GRanges("chr1", IRanges(10e6,10.1e6))
sce[sce %over% x,]
```

## Add cell annotations

Cell annotations were already generated using *Seurat* [@seurat]. The
script is saved in this package in `inst/scripts/seurat.R`. Here we
will use them for size factor estimation and for visualization.

```{r}
ids <- readRDS(file.path(extdata, "idents.rds"))
top10 <- read.csv(file.path(extdata, "top10.csv"))
```

We subset to the cells that we have IDs for, and attach the labels in
the correct order:

```{r}
idx <- colnames(sce) %in% names(ids)
table(idx)
sce <- sce[,idx]
sce$cluster <- ids[colnames(sce)]
```

The number of cells per cluster:

```{r}
table(sce$cluster)
```

Note that the different clusters have different total counts, for
example:

```{r}
cs <- colSums(assays(sce)[["counts"]])
# cells with higher number of UMI
more.umi <- cs > 10000
(tab <- table(more.umi, sce$cluster))
100 * round(prop.table(tab,2),2) # percent
# cell with lower number of UMI
fewer.umi <- cs < 5000
(tab <- table(fewer.umi, sce$cluster))
100 * round(prop.table(tab,2),2) # percent
```

We have a data.frame with the top marker genes per cluster, as
identified with *Seurat*.

```{r}
head(top10)
```

## Plotting counts with uncertainty

In this section, we will begin to plot the counts for cells, for
specific genes, and showing the *inferential uncertainty* as
quantified by *alevin*. This is a unique aspect to the *alevin*
quantification method, that it retains gene multi-mapping reads,
instead of discarding these reads. *alevin* can also attach a measure
of uncertainty to each count in the matrix. *alevin* computes the mean
and variance of *inferential replicates* which are generated by
bootstrapping the read data. `tximeta` will import these inferential
mean and variance matrices by default (they are also sparse, similar
to the counts matrix).

We will first visualize the uncertainty, and later give an
example of a set of genes where the uncertainty is indicating
significant quantification uncertainty which is useful to consider
when performing EDA (exploratory data analysis) or statistical
analysis.

For a later demonstration of scaling, we will sort the cells by the
total count (this is not something you would necessarily do in a
typical analysis).

```{r}
o <- order(colSums(assays(sce)[["counts"]]), decreasing=TRUE)
sce <- sce[,o]
```

We can now use a plotting function `plotInfReps` from the *fishpond*
package in order to plot the inferential mean and variance for the
cells, for various genes.

```{r plot-basic}
library(fishpond)
plotInfReps(sce, idx="ENSG00000167286.9",
            x="cluster", mainCol="SYMBOL",
            legend=TRUE)
```

Here, we have `x` as a grouping variable for coloring the points, one
can also specify `x` to be a continuous covariate,
e.g. pseudo-time. An additional covariate `cov` can be used for
additional grouping, e.g. batches, donors or lineages.

Here we show the same plot but now subsetting the number of cells: 

```{r plot-medium}
idx <- sample(ncol(sce),200)
plotInfReps(sce[,idx],
            idx="ENSG00000167286.9",
            x="cluster", mainCol="SYMBOL",
            legend=TRUE)
```

This time only 100 cells:

```{r plot-small}
idx <- sample(ncol(sce),100)
plotInfReps(sce[,idx],
            idx="ENSG00000167286.9",
            x="cluster", mainCol="SYMBOL",
            legend=TRUE)
```

The plots change their features across different sample sizes for ease
of viewing the uncertainty of quantification for individual cells.

We can also choose to plot the cells in their original order (the
default for scRNA-seq is to sort by the mean value):

```{r plot-no-order}
plotInfReps(sce, idx="ENSG00000167286.9",
            x="cluster", mainCol="SYMBOL",
            reorder=FALSE)
```

## Scaling with size factors

We use `computeSumFactors` [@Lun2016] from the *scran* package
[@scran] to compute size factors that are stored in
`sizeFactors(sce)`. 

```{r size-factors}
library(scran)
sce <- computeSumFactors(sce, clusters=sce$cluster)
plot(cs, sizeFactors(sce), xlab="column sum", ylab="sum factor")
```

Now we demonstrate un-scaled counts and counts scaling with size 
factors just computed. (Note in the second plot that the first cell in
each group isn't the highest anymore.)

```{r scaling}
par(mfrow=c(2,1), mar=c(2.5,4.5,1,1))
plotInfReps(sce, idx="ENSG00000167286.9",
            x="cluster", main="",
            reorder=FALSE)
plotInfReps(sce, idx="ENSG00000167286.9",
            x="cluster", main="",
            applySF=TRUE, reorder=FALSE)
```

## Inferential variance (uncertainty)

In this second to last section, we will briefly talk about how the
inferential uncertainty as stored in the variance assay may be useful
in practice.

Many users may prefer to just work with the counts matrix, and not
consider the inferential mean and variance. We have found that,
globally, this may not lead to too much of a problem, but for certain
genes, it may be important to use the inferential variance in cases
where it may signal difficult to quantify genes. Inferential
uncertainty for bulk and single cell RNA-seq differential expression
was a focus of the *Swish* nonparametric statistical method [@swish]
which we do not demonstrate here, but one can refer to
the vignette of [swish](https://bioconductor.org/packages/swish) for
more details (in particular the section on *alevin* data).

Let's start by visualizing the uncertainty across all values in the
matrix: 

```{r}
var <- as.vector(assays(sce)[["variance"]])
mu <- as.vector(assays(sce)[["mean"]])
idx <- mu > 3
df <- data.frame(log10mean=log10(mu[idx]),
                 log10var=log10(var[idx]))
```

```{r var-mean, fig.height=4}
library(ggplot2)
ggplot(df, aes(log10mean, log10var)) +
  geom_hex(bins=100) + 
  geom_abline(intercept=0, slope=1, col="red")
```

The values around the red line indicate not much uncertainty
(bootstrapping reads gives us roughly multinomial, and so also
approximately Poisson, variability around the mean count). However we
can see a tail of higher uncertainty values in the matrix, where the
inferential variance is for example, up to 10 times higher than the
mean. 

We can also plot this summarized to a single value per gene, here we
calculate the 99% quantile of the ratio of variance over mean, per
gene, and plot this over the mean:

```{r}
library(matrixStats)
rratio <- rowQuantiles(as.matrix(
  assays(sce)[["variance"]] /
  (assays(sce)[["mean"]] + 1)), probs=.99)
rmu <- rowMeans(assays(sce)[["mean"]])
idx <- rmu > .1
df <- data.frame(log10mean=log10(rmu[idx]),
                 rratio=rratio[idx],
                 gene=mcols(sce)$SYMBOL[idx])
```

A number of genes have a ratio above 2, but then two genes in this
dataset stand out above the rest, and they happen to be a ribosomal
gene, and another gene that is a read-through of the same locus:

```{r high-var-mean-ratio}
with(df, plot(log10mean, rratio))
high.uncert <- which(df$rratio > 10)
with(df[high.uncert,],
     points(log10mean, rratio, pch=20, col="red"))
with(df[high.uncert,],
     text(log10mean, rratio, gene, pos=c(2,4)))
mcols(sce)$SYMBOL[idx][high.uncert]
```

Some of the counts for the ribosomal gene and its read-through:

```{r plot-high-uncert}
par(mfrow=c(2,1), mar=c(2.5,4.5,1,1))
plotInfReps(sce[,1:100], idx="ENSG00000265681.7",
            x="cluster", mainCol="SYMBOL")
plotInfReps(sce[,1:100], idx="ENSG00000215472.10",
            x="cluster", mainCol="SYMBOL")
```

The extra uncertainty on the counts for this gene indicate that
*alevin* was not certain if the reads should come from the gene or its
read-through. While in this case, the ribosomal gene may not be over
interest, there are other cases (e.g. developmental or immune genes
with high sequence homology) where information of the uncertainty of
quantification can be useful in interpreting the data.

## Downstream analysis with Seurat

As we previously showed how to construct a *SingleCellExperiment*
which can be used with other Bioconductor workflows, we also
demonstrate how it is easy to convert the `sce` object into an object
for use with the *Seurat* R package [@seurat] for single cell
analysis. As we noted, *Seurat* was already used to identify the cell
types (with the script stored in `inst/scripts/seurat.R`).

We now load *Seurat* and create a *Seurat* object:

```{r}
library(Seurat)
cts <- assays(sce)[["counts"]]
pbmc <- CreateSeuratObject(cts)
```

We can easily create violin plots, for example:

```{r seurat-violin}
mt.genes <- rownames(sce)[as.logical(seqnames(sce) == "chrM")]
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, features=mt.genes)
feats <- c("nFeature_RNA", "nCount_RNA", "percent.mt")
VlnPlot(pbmc, features=feats, ncol=3)
```

From this point, one can use the `pbmc` object for use in *Seurat*
workflows, for example,
the [vignettes](https://satijalab.org/seurat/vignettes.html) on the
*Seurat* website. 

## References
